# Discretisation {#sec-discretisation}

Geographers, maybe more than others because they like to produce maps, are often tempted to cut a numerical vector into a set classes.

A typical use is the cartography of a continuous variable in a set of 5, 6 or 7 groups, knowing the human eye has difficulties to disentangle more colours. GIS and mapping software all have a menu where "discretisation" is made using a number of manually defined limits of preset algorithms, such as "Natural breaks", "quantiles", etc.

In R, the function `cut()`\` is the base function to divide a range of numeric values x into intervals by a number of break values. It outputs a new factor where each value x is given a level according to which interval they fall. The breaks are either

-   a scalar greater or equal to 2 for the range to be into equal length pieces.

-   a set of breaks to be used as the upper and lower limits of each discrete category

```{r}
x<-rnorm(20, mean = 100, sd=5)
xf4<-cut(x, breaks=4)
head(xf4)
table(xf4)
```

Notice how the label clearly indicates the (default) closing on the right of each interval

```{r}
x<-rnorm(20, mean = 100, sd=5)
xf6<-cut(x, breaks=c(min(x),-90,95,100,105,110, max(x)))
head(xf6)
table(xf6)
```

Since we can choose any breaks, it is pretty easy to adapt and use any discretisation method one would find elsewhere, e.g. in mapping packages.

There is a wonderful package, `classint`, that does so and where you can simply choose the discretisation methodology.

Let's explore! \
<https://cran.r-project.org/web/packages/classInt/classInt.pdf>
