## Introduction {.smaller}

ggplot2 is an **extension of the tidyverse** that enables the generation of charts with a consistent and powerful syntax. It requires learning an additional **'mini-language'** but allows for the efficient construction of complex graphs.

#### Library

Here, we simply load the various packages that will allow us to **manipulate data tables** and **create figures**:

-   `tidyverse` is a package that contains several packages designed for data manipulation in R (stringr, dplyr, ggplot2)

-   `ggplot2` which allows you to create **complex graphics** and **observing data** with visuals

In **ggplot2**, a **geom** (short for “geometric object”) is a layer that defines how data points are visually represented in a plot. Each type of plot, such as points, lines, bars, etc., is created using a specific geom function. Here are some common examples:

-   **`geom_point()`**: for scatter plots.

-   **`geom_bar()`**: for bar charts.

-   **`geom_histogram()`**: for histograms.

-   **`geom_boxplot()`**: for box plots.

These functions are added to a basic ggplot object using the `+` operator.

Example of code :

ggplot() +

geom\_............(data, aes(x = x, y = y, fill = color)) +

labs(title = "title", x = "abs", y = "ord") +

theme_bw()

```{r, warning = FALSE}
#| label: Library
#| #| label: load-packages

library(ggplot2)
library(questionr) # insee data
library(dplyr) 
library(RColorBrewer)
library(readr) 
library(viridis)
library(tidyr)

display.brewer.all()
```

## Data {.smaller}

We will present different code examples with one dataset, but the codes are applicable to all datasets. Therefore, it is important to save our codes properly so that we can copy them for other data. Here is the dataset :

The file **'rp2018'** contains, for all the municipalities in France in 2018, the following variables will be use.

-   pop_tot (population)

-   etud (student)

-   cadres (senior executive)

-   locataire (tenant)

```{r}

data(rp2018)
head(rp2018)

```

## Basic plot - ggplot {.smaller}

Here, we indicate to R that we want to plot a scatterplot, so we need to specify a "x" and “y” for our graph. Here the example of the basic R plot, and the ggplot plot.

```{r}

plot(rp2018$etud, rp2018$cadres) # plot(x, y)

ggplot() + 
  geom_point(data = rp2018, aes(x = etud, y = cadres))

```

## Histogram {.smaller}

Usually, when you're interested in a variable, you look at its distribution. First, a classic basic histogram: you specify the data table in the `ggplot()` function, and the variable of the table that will be used to create the plot in the `aes()` function. For example, we need to specify which column of the table will be represented on the `x-axis`, on the `y-axis`, or which column will allow us to color certain elements or modify the shape of the points, for instance. We then add successive layers to the graph using the `+` sign to achieve the desired rendering.

Here, we indicate to R that we want to plot a histogram, so we don’t need to specify a “y” for our graph:

```{r}

ggplot(rp2018, aes(x = etud)) +
  geom_histogram()

```

The problem is that the figure is quite unattractive, isn’t it? So, we will modify it with options: `fill` for the bar fill, `color` for the border, and `alpha` for transparency. The theme allows us to change the overall appearance of the figure. Here you can see all theme : <https://ggplot2.tidyverse.org/reference/ggtheme.html>

Below, `geom_histogram()` is used to plot a histogram.

-   We can also modify the number of classes represented by the histogram with the `bins` option.

-   To modify the size of the classes represented by the histogram, we use the `binwidth` option.

Try both codes below:

```{r}


ggplot(rp2018, aes(x = etud)) +
  geom_histogram(fill = "red",
                 color = "black",
                 binwidth = 5, # reduce or increase the width of the bar (number of classes)
                 alpha = 0.5) + 
  labs(title = "Student proportion in France per municipalities", 
       x = "Student (%)", 
       y = "Count") + 
  theme_bw() 


```

## Density {.smaller}

`geom_density()` is used to plot the density curve.

```{r}


ggplot(rp2018, aes(x = etud)) +
  geom_density(fill = "red",
                 color = "black",
                 alpha = 0.5) + 
  labs(title = "Student proportion in France per municipalities", 
       x = "Student (%)", 
       y = "Count") + 
  theme_bw() 
```

## QQplot {.smaller}

A `QQ plot` (quantile-quantile plot) indicates a normal distribution when the points closely follow a straight line, suggesting that the quantiles of the data match the quantiles of a **normal distribution**. A normal Q-Q plot is a common application where one set of data is plotted against a theoretical normal distribution. If the points form a straight line, the data can be considered to be normally distributed. Here is a good QQ-plot reference : <https://www.tjmahr.com/quantile-quantile-plots-from-scratch/>

```{r}

ggplot(rp2018, aes(sample = etud)) +
  geom_qq(color = "red") +
  labs(x = "Student (%)") +
  theme_bw()
```

## Boxplot {.smaller}

Let's look at the yields in the form of **boxplots** for each variety. To do this, we use the `geom_boxplot()` function. Please note that the color of boxplots and barplots is managed not with `color` but with `fill` (points and lines are managed with 'color'). Other palettes are available at the following address: <https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html>

```{r}


ggplot(data = rp2018, aes(x = code_region, y = etud, fill = region)) +
  geom_boxplot() +
  labs(title = "Proportion of cadre per region", x = "Region", y = "Student(%)", fill = "Region") + 
  scale_fill_brewer(palette = "Set1") + # color
  theme_bw() 


```

The `colorRampPalette()` is a the function that returns you a fixed number of colors that interpolate the palette. Again we have a function pal() that was returned by colorRampPalette(), this time interpolating a palette containing colors. But now, the pal() function takes an integer argument specifing the number of interpolated colors to return.

The reference for the packages legocolors : <https://cran.r-project.org/web/packages/legocolors/readme/README.html>

```{r, message = FALSE, }


pal1 <- colorRampPalette(c("lightblue", "purple", "red", "green"))
pal2 <- colorRampPalette(brewer.pal(9, "Set1")) # 9 to use all the color
library(legocolors)
pal3 <- colorRampPalette(legoCols$hex[2:13])

# length(rp2018$code_region)
# length(unique(rp2018$code_region))

ggplot(data = rp2018, aes(x = code_region, y = etud, fill = region)) +
  geom_boxplot() +
  labs(title = "Proportion of cadre per region", x = "Region", y = "Student(%)", fill = "Region") + 
  scale_fill_manual(values = pal2(17)) + # you can change the pal() here (see above)
  theme_bw()
```

## Multi-boxplot {.smaller}

We could also put the blocks in x, and color according to the variable. It allows to make a graphics with 3 dimensions.

For the **'rp2018'** data, we can create categories based on regions or departments; however, this still results in a large number. In the example below, I still made the boxplot graph but removed the legend. In this case, it is still better to do a multiple plots with a grid format.

```{r}

ggplot(rp2018, aes(x = region, y = etud, fill = code_departement)) +
  geom_boxplot(outlier.shape = NA) + 
  scale_fill_manual(values = pal2(100)) + 
  labs(title = "Department in the region", x = "Region", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# 44 Grand-Est, 76 Occitanie, 53 Bretagne, 28 Normandie

rp2018_reg44 <- rp2018[rp2018$code_region == "44",]

g44 <- ggplot(rp2018_reg44, aes(x = code_departement, y = etud, fill = code_departement)) +
  geom_boxplot() + 
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Grand Est", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none")

rp2018_reg76 <- rp2018[rp2018$code_region == "76",]

g76 <- ggplot(rp2018_reg76, aes(x = code_departement, y = etud, fill = code_departement)) +
  geom_boxplot() + 
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Occitanie", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none")

rp2018_reg53 <- rp2018[rp2018$code_region == "53",]

g53 <- ggplot(rp2018_reg53, aes(x = code_departement, y = etud, fill = code_departement)) +
  geom_boxplot() + 
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Bretagne", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none")

rp2018_reg28 <- rp2018[rp2018$code_region == "28",]

g28 <- ggplot(rp2018_reg28, aes(x = code_departement, y = etud, fill = code_departement)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Normandie", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none")

library(gridExtra)
grid.arrange(g44, g76, g53, g28,
          nrow = 2, ncol = 2)
```

## Boxplot with point {.smaller}

To display points on a boxplot, you can use the `geom_jitter()` function, which displays the points by offsetting them from each other to avoid overlap. When displaying the points, you should add the `'outlier.shape = NA'` option in the `geom_boxplot()`function to prevent displaying the same points twice.

```{r}

rp2018_reg44 <- rp2018[rp2018$code_region == "44",]

ggplot(rp2018_reg44, aes(x = departement, y = etud, fill = departement)) +
  geom_boxplot(outlier.shape = NA) + #exclude point outlier
  geom_jitter(width = 0.1, color = "black") + # jitter allows to arrange the place of the point comparaing to the geom_point
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Grand Est", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

## Violinplot {.smaller}

One last one, the violin plot, which shows us the distribution curve. We use the `geom_violin()` function for this.

```{r}

ggplot(rp2018_reg28, aes(x = code_departement, y = etud, fill = code_departement)) +
  geom_violin() +
  geom_boxplot(width=0.1, color="grey", alpha=0.2) +
  labs(title = "Normandie", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none")

#choose your color with a vector of colors

ggplot(rp2018_reg28, aes(x = code_departement, y = etud, fill = code_departement)) +
  geom_violin(bw = 2) +
  geom_boxplot(width=0.1, color="grey", alpha=0.2) +
  scale_fill_manual(values = c("black", "red", "yellow", "darkgreen", "pink"),
                    limits = c("14", "27", "50", "61", "76")) + # color
  labs(title = "Normandie", x = "Department", y = "Student(%)") +
  theme_bw() +
  theme(legend.position = "none")
```

## Scatterplot {.smaller}

We can also create a scatterplot (a graph with points) between **higher diplomas** and **cadres** for example. The `geom_point()` option allows us to display points (with continuous data, for instance).

The `scale_color_gradient()` here allows us to directly detect the numerical values of the variables and produce a continuous color palette for data visualization. This allows us to integrate another variable on the points of the graph, and we can also apply it to the size to compare the x-axis and y-axis of the data to find the effect of other variables. For example, we can see that the higher the proportion of students, the more apartment rentals we observe. The `scale_color_gradientn()`here allows to

```{r}

ggplot(rp2018, aes(x = dipl_sup, y = cadres)) + 
  geom_point() + 
  theme_classic()


ggplot(rp2018, aes(x = dipl_sup, y = cadres, color = chom, size = pop_tot)) + 
  geom_point() + 
  scale_color_gradient("Unemployment (%)", low = "white", high = "blue") +
  scale_size(range = c(0, 10)) +
  theme_classic()


ggplot(rp2018, aes(x = dipl_sup, y = cadres, color = chom, size = pop_tot)) +  
  geom_point() + 
  scale_color_gradientn(name = "Unemployment (%)",
                        colors = c("white", "lightblue", "blue", "black"), 
                        limits = c(min(rp2018$age_15_29), max(rp2018$age_15_29))) +
  scale_size(range = c(0, 10)) +
  theme_classic()
```

## Scatterplot and regression line {.smaller}

We can also try to put the color of the points according to the variable (to color points, use `color`), and we add regression lines with the `geom_smooth()` for the **higher diplomas** and the **cadres**.

```{r}


ggplot(rp2018, aes(x = dipl_sup, y = cadres, color = chom)) +  
  geom_point() + 
  geom_smooth(method = "lm", se = F) +
  scale_color_gradientn(name = "Unemployment (%)",
                        colors = c("white", "lightblue", "blue", "black"), 
                        limits = c(min(rp2018$age_15_29), max(rp2018$age_15_29))) +
  scale_size(range = c(0, 10)) +
  theme_classic()


# for numerical variables, you can also you the color with the viridis package
ggplot(rp2018, aes(x = dipl_sup, y = cadres, color = chom)) +  
  geom_point() + 
  geom_smooth(method = "lm", se = F) +
  scale_color_viridis(name = "Unemployment (%)") +
  theme_classic()
```

## Barplot {.smaller}

Here’s a quick example to understand the proportion of forest area in each canton of Luxembourg:

For instance, you can create a bar chart where the x-axis represents the cantons and the y-axis represents the proportion of forest area. Using a function like `geom_bar()` or `geom_col()` to create a barplot.

To specify that we want the height of the barplot to represent the value indicated in our data frame, we need to add `stat = "identity` in the `geom_bar()` function. Normally, `geom_bar()` counts the number of cases at each x position (**as an histogram**), but with `stat = "identity"`, it uses the values in the data directly. The **'width'** option allows you to adjust the width of the barplot.

Source : <https://ggplot2.tidyverse.org/reference/geom_bar.html>

```{r}
forest_area <- read.csv2("C:/Users/leandre.fabri/Documents/R/MAGEO_06.41/data/statec/forest_area_canton.csv", sep = ",") 


ggplot(forest_area, aes(x = GEO..Géographie, y = OBS_VALUE, fill = GEO..Géographie )) +
  geom_bar(stat = "identity",
           width = 0.5) + # width of the bar
  theme_bw() +
  labs(title = "Area of the afforestation rate", x = "Cantons", y = "Afforestation rate") +
  scale_fill_manual(values = pal2(17)) +
  theme(legend.position = "none") 


ggplot(forest_area, aes(x = GEO..Géographie, y = OBS_VALUE, fill = GEO..Géographie )) +
  geom_bar(stat = "identity",
           width = 0.5) + # width of the bar
  theme_bw() +
  labs(title = "Area of the afforestation rate", x = "Cantons", y = "Afforestation rate") +
  scale_fill_manual(values = pal2(17)) +
  theme(legend.position = "none") +
  coord_flip() # allow to keep the real name of the canton

```

## To help you {.smaller}

[data-to-viz](https://www.data-to-viz.com/)

[r-graph-gallery](https://r-graph-gallery.com/)

[cookbook-r](http://www.cookbook-r.com/Graphs/ "cookbook-r")

[juba.github](https://juba.github.io/tidyverse/08-ggplot2.html)

#### **Data of the classes** :

library(questionr) -\> rp2018

Forest area by canton and commune -\> [lustat.statec.lu](https://lustat.statec.lu/vis?fs%5B0%5D=Th%C3%A8mes%2C1%7CTerritoire%20environnement%20et%20%C3%A9nergie%23A%23%7CEnvironnement%23A3%23&pg=20&fc=Th%C3%A8mes&df%5Bds%5D=ds-release&df%5Bid%5D=DF_X038&df%5Bag%5D=LU1&df%5Bvs%5D=1.0&pd=2010%2C2010&dq=..A)

## Your turn! {.smaller}

-   Go take the data on land use proportions for some cities in the ouest of Europe : `land_use_EU_ouest.rds`

-   Create a few plots with a '**ggplot2**' style.

The data comes from the **urban atlas 2018** shows the proportion and the area of different classes of land use per city.

-   **p\_**.... is the proportion in %

-   **area\_**..... is the surface in m²

-   **pop_tot** of the city

-   **perimeter_tot** and **area_tot** of the city
