# Usual Continuous Distributions

In addition to the normal and uniform distributions, a number of other continuous distributions are often used, corresponding to different assumptions about the processes that shape the population.

## Exponential

A random variable $X \in [0;+\infty[$ following an exponential distribution with parameter $\lambda > 0$ can be expressed as: $X \sim \epsilon(\lambda)$ with expectation $E(X) = 1/ \lambda$ and variance $Var(X) = 1/ \lambda^2$.

Its **density function** can be written as:

$$f(x) = λ {e}^{- λ x} \text{ for } x \geq 0$$ $$f(x) = 0 \text{ for } x < 0$$.

Its **cumulative distribution function** is defined by:

$$F_X(x) = 1 - \exp{(- \lambda x)} \text{ for } x \geq 0$$ $$F_X(x) = 0 \text{ for } x < 0$$

```{r exp}
x <- seq(0,8,0.01)
dExp.5 <- dexp(x, .5)
dExp1 <- dexp(x, 1)
dExp2 <- dexp(x, 2)
dExp3 <- dexp(x, 3)
```

```{r expDensity}
plot(x, dExp3, type = 'l', col = 'green',
     main = 'Density function of X~Exp(lambda)', ylab = 'f(x)')
lines(x, dExp2, col = 'gold')
lines(x, dExp1, col = 'darkorange')
lines(x, dExp.5, col = 2)
legend("topright", lty = 1,
       legend = c("lambda = 0.5", "lambda = 1", 
                  "lambda = 2", "lambda = 3"),
       col = c(2, 'darkorange', 'gold', 'green'))
```

```{r expCDF}
plot(x, pexp(x, 3), type = 'l', col = 'green',
     main = 'Cumulative Distribution Function of X~Exp(lambda)', ylab = 'F(x)')
lines(x, pexp(x, 2), col='gold')
lines(x, pexp(x, 1), col='darkorange')
lines(x, pexp(x, .5), col=2)
legend("bottomright", lty = 1,
       legend = c("lambda = 0.5", "lambda = 1", "lambda = 2", "lambda = 3"),
       col = c(2, 'darkorange', 'gold', 'green'))
```

## Chi-square

A random variable $X$ following a chi-square (or khi-square) distribution can be expressed as: $X \sim \chi^2(p)$ with parameter $p > 0$, expectation $E(X) = p$, variance $Var(X) = 2p$.

Its **density function** is

$$f_X(x) = \dfrac {1} {2^{n/2} \gamma(n/2)} x^{(n/2)-1} \exp{(\dfrac{-x}{2})} \text{ for } x \geq 0$$

$$f_X(x) = 0 \text{ for } x < 0$$

```{r chi2}
x <- seq(0,8,0.01)
dChiSq1 <- dchisq(x, 1)
dChiSq2 <- dchisq(x, 2)
dChiSq3 <- dchisq(x, 3)
dChiSq6 <- dchisq(x, 6)
```

```{r chi2Density}
plot(x, dChiSq1, type = 'l', col = 2, ylim = c(0, 1.25),
     main = 'Density function of X~Chi^2(p)', ylab = 'f(x)')
lines(x, dChiSq2, col = 'darkorange')
lines(x, dChiSq3, col = 'gold')
lines(x, dChiSq6, col = 'green')
legend("topright", lty = 1,
       legend = c("p = 1", "p = 2", "p = 3", "p = 6"),
       col = c(2, 'darkorange', 'gold', 'green'))
```

```{r chi2CDF}
plot(x, pchisq(x, 1), type = 'l', col = 2,
     main = 'Cumulative Distribution Function of X~Chi2(p)', ylab = 'F(x)')
lines(x, pchisq(x, 2), col='darkorange')
lines(x, pchisq(x, 3), col='gold')
lines(x, pchisq(x, 6), col='green')
legend("bottomright", lty = 1,
       legend = c("p = 1", "p = 2", "p = 3", "p = 6"),
       col = c(2, 'darkorange', 'gold', 'green'))
```

## Student t-dsitribution

A random variable $X$ following a student distribution, can be expressed as: $X \sim t(n)$ with parameter $n$ degrees of freedom.

Its **density function**is

$$f(x) = \dfrac{1}{\sqrt{n\pi}} \dfrac {\gamma{(\dfrac{n+1}{2})}} {\gamma{(\dfrac{n}{2})}} \dfrac {1} {(1+\dfrac{x^2}{n})^{(n+1)/2}} \text{ for all } x\in R$$ We have: $$E(X) = 0 \text{ for } n \ge 2$$ $$Var(X) = \dfrac{n}{n-2} \text{ for } n \ge 3$$

If we have a random variable $U \sim N (0, 1)$ and a random variable $X \sim \chi^2(n)$ which are independent, then the random variable $T_n = \dfrac{U}{\sqrt{X/n}}$ follows a student distribution $t(n)$

```{r student}
x <- seq(-8, 8, 0.01)
dStudent1 <- dt(x, 1)
dStudent2 <- dt(x, 2)
dStudent4 <- dt(x, 4)
```

```{r studentDensity}
plot(x, dNorm01, type = 'l', col = 'darkgray', xlim = c(-4, 4), lty = 3,
     main = 'Density function of X~t(n)', ylab = 'f(x)')
lines(x, dStudent4, col = 'gold')
lines(x, dStudent2, col = 'darkorange')
lines(x, dStudent1, col = 2)
legend("topright", lty = c(rep(1, 3), 3),
       legend = c("n = 1", "n = 2", "n = 4", "N(0,1)"),
       col = c(2, 'darkorange', 'gold', 'darkgray'))
```

```{r studentCDF}
plot(x, pnorm(x), type = 'l', col = 'darkgray', lty = 3,
     main = 'Cumulative Distribution Function of X~t(n)', ylab = 'F(x)')
lines(x, pt(x, 2), col='gold')
lines(x, pt(x, 3), col='darkorange')
lines(x, pt(x, 6), col=2)
legend("bottomright", lty = c(rep(1, 3), 3),
       legend = c("n = 1", "n = 2", "n = 4", "N(0,1)"),
       col = c(2, 'darkorange', 'gold', 'darkgray'))
```

## Fisher-Snedecor

A random variable $X$ following a Fisher-Snedecor distribution can be expressed as $X \sim F(n, p)$ with parameters $n$ and $p$ degrees of freedom.

Its **density function** is

$$f_X(x) = \dfrac {\gamma(\dfrac{n+p}{2})} {\gamma(\dfrac{n}{2}) \gamma(\dfrac{p}{2})} (\dfrac{n}{p})^{(n/2)} \dfrac{x^{\dfrac{n-2}{2}}}{(1 + \dfrac{n}{p}x)^{\dfrac{n-2}{2}}} \text{ for } x \geq 0$$ $$f_X(x) = 0 \text{ for } x < 0$$

We have:

$$E(X) = \dfrac{p}{p-2} \text{ for } p \ge 3$$ $$Var(X) = \dfrac{2p^2(n+p-2)}{n(p-2)^2(p-4)} \text{ for } p \ge 5$$

If we have two random variables $X$ and $Y$ independent such that $X \sim \chi^2(n)$ and $Y \sim \chi^2(p)$, then the random variable $\dfrac{X/n}{Y/p}$ follows a Fisher distribution $F(n, p)$

```{r fisher}
x <- seq(0, 3, 0.01)
dF1 <- df(x, 1, 1)
dF10 <- df(x, 10, 10)
dF20 <- df(x, 20, 20)
dF50 <- df(x, 50, 50)
```

```{r fisherDensity}
plot(x, dF1, type = 'l', col = 2, 
     main = 'Density function of X ~ F(n, p)', ylab = 'f(x)')
lines(x, dF10, col = 'darkorange')
lines(x, dF20, col = 'gold')
lines(x, dF50, col = 'green')
legend("topright", lty = 1,
       legend = c("n = p = 1", "n = p = 10", "n = p = 20", "n = p = 50"),
       col = c(2, 'darkorange', 'gold', 'green'))
```

```{r fisherCDF}
plot(x, pf(x, 50, 50), type = 'l', col = 'green',
     main = 'Cumulative Distribution Function of X ~ F(n, p)', ylab = 'F(x)')
lines(x, pf(x, 20, 20), col='gold')
lines(x, pf(x, 10, 10), col='darkorange')
lines(x, pf(x, 1, 1), col=2)
legend("bottomright", lty = 1,
       legend = c("n = p = 1", "n = p = 10", "n = p = 20", "n = p = 50"),
       col = c(2, 'darkorange', 'gold', 'green'))
```
